Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-19 ThesisJP/Background/background.tex
Started writing compare cpu and gpu.

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see why there is a big difference in peak performance for a CPU and GPU.

@@

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
\label{fig:gpu_grid}
\end{figure}

>>>>

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and GPU.

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD Extensions}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced Vector Extensions}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers which means that eight 32 bit floating point values can be processed per clock cycle. This extension is found in the latest CPUs from Intel. AVX-512 extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest quad-core CPU from Intel, the Haswell architecture, can therefore process 32 32 bit floating point values per clock cycle. With a clock frequency of 3.9 GHz this sums up to a theoretical peak single precision performance of 124.8 giga single precision floating point operations per second (GFLOPS). In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing 16 32 bit floating point values at 1.1 GHz. The total theoretical peak single precision FLOPS for a high-end Haswell CPU is therfore $124.8 + (40*16*1.1) = 828.8$ GFLOPS as visualized in Fig\,\ref{fig:cpu_vs_gpu}.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors. The Fermi GPU architecture by Nvidia, which is used the most in this theses, have up to 16 SIMD units capable of performing 32 32 bit floats per clock cycle at 772 MHz. The theoretical peak single precision FLOPS for this architecture is by that found to be $16*32*772 = $ (needs to add FMA and SP-units, What about FMA on the CPU?)

@@

In this section we have focused on CPUs from Intel and GPUs from Nvidia, note that GPUs and CPUs from AMD have similar features.

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an element-wise matrix square.}
\label{fig:gpu_grid}
\end{figure}
